Geometric Filters for Metric Spaces

### 2018-11-28 paper/draft.tex
add Michaels proof

\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{comment}
\usepackage{braket}

@@

The algorithm that we analyze can be roughly described as follows. 
Pick a random point $p_i$,
compute $\dist(p_i, q)$ and use this distance combined with pairwise distances $\dist(p_i, p_j)$
to update lower bounds for other distances $\dist(p_j, q)$.
If we see that the lower bound for $\dist(p_j, q)$ is larger than $\dist(p_i, q) / (1+\eps)$, then 
we can discard the point $p_j$. If all points were discarded, we found an approximate nearest neighbor;
otherwise pick a random point $p_j$ from the remaining ones and compute $\dist(p_j, q)$.
There are two possibilities: if $\dist(p_j, q) \leq \dist(p_i, q) / (1+\eps)$,
then we choose $p_j$ as the current candidate for an approximate nearest neighbor,
otherwise we keep $p_i$ in this role. In any case, the new computed distance
may also allow us to discard more than one point, as figure \ref{fig:discarded_balls} shows. In
this figure points inside the shaded smaller ball cannot be a better candidate,
and the same is true for the points in the shaded area outside of the larger ball.
The precise radii of these balls are given in the following lemma.

\begin{lemma}
    Let $p_i, p_j \in P$ be such that $\dist(p_j, q) \geq \frac{\dist(p_i, q)}{1 + \eps}$,
    and denote $r_i = \dist(p_i, q)$, $r_j = \dist(p_j, q)$.
    Then for every $x \in B(p_j, r_j - r_i / (1+\eps))$ we have $\dist(x, q) \geq r$
    and for every $y \notin B(p_j, r_j + r_i / (1+ \eps))$ we have $\dist(y, q) \geq r$,
    so all points of $P$ inside the smaller ball and all points of $P$ outside the larger ball
    can be discarded.
\end{lemma}
\begin{proof}
    The triangle inequality.
\end{proof}

@@

\begin{algorithmic}
\Procedure{Discard\_1}{$L, p_i$}
    \State {$r_i = \dist(q, p_i)$}
    \ForAll { $p_j \in L$}
       \If {  $ \dist(p_i, p_j) > r_i \left(1 + \frac{1}{ 1 + \eps} \right)$ }
           \State {Remove $p_j$ from $L$}
       \EndIf
    \EndFor
\EndProcedure
\end{algorithmic}

\begin{algorithmic}
\Procedure{Discard\_2}{$L, w, c$}
    \State {$r_w = \dist(q, w)$}
    \State {$r_c = \dist(q, w)$}
    \Comment {Distances $r_w$ and $r_c$ are cached}
    \ForAll { $p_j \in L$}
       \If { $ \dist(w, p_j) < r_w - r_c / ( 1 + \eps) $ }
           \State {Remove $p_j$ from $L$}
           \Comment {In particular, the worse candidate $w$ will be removed here}
       \EndIf
       \If {  $ \dist(w, p_j) > r_w + r_c / ( 1 + \eps) $ }
           \State {Remove $p_j$ from $L$}
       \EndIf
    \EndFor
\EndProcedure
\end{algorithmic}

\begin{algorithmic}
\label{alg:ann_blind}
\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
    \State {Initialize array $L$ with a random permutation of $P$}
    \State {$ c \gets \mbox{the first element of }L$}
    \State {Compute $r_i = \dist(q, p_i)$}
    \State \Call{Discard\_1}{$L, c$}
    \While {$|L| > 1$}
        \ForAll {$p_j \in L $}
            \State {Compute $r_j = \dist(p_j, q)$}
            \Comment {Set the worse candidate $w$ to be either $p_i$ or $p_j$}
            \If {$r_j < r_i$ }
                \State {$ c \gets p_j $}
                \State \Call{Discard\_1}{$L, c$}
                \State {$ w \gets p_i $}
            \Else
                \State {$ w \gets p_i $}
            \EndIf
            \State \Call{Discard\_2}{$L, w, c$}
        \EndFor
    \EndWhile
    \State \Return {the unique element left in $L$}
\EndFunction
\end{algorithmic}

@@

The proof is based on the following lemma.
\begin{lemma}
    There exists a constant $C$ that depends only on $\eps$ and the doubling dimension
    such that after computing at most $C$ distances $\dist(p_k, q)$
    either the algorithm verifies that the current candidate $p_i$ is a correct answer and terminates
    or the algorithm finds another candidate $p_j$ such that $r_j < r_i$.
\end{lemma}
\begin{proof}
    First, all points that are not contained in the ball $B(p_i, 2r_i)$
    are discarded without computing the distance to them.
    Now we have to examine all points $p_j$ inside this ball. 
    Suppose that we cannot discard the current point $p_j$ based
    on the lower bound for $r_j$, then we must compute the distance $r_j$.
    There are two possibilities: either $r_j \leq r_i$ or $r_j > r_i$.
    The latter case means that we discard the point $p_j$,
    but we also discard all points of $P$ that are contained in some ball around it.
    The triangle inequality tells us that we can take the ball of radius $r_j - r_i / (1 + \eps)$,
    but in this proof it is sufficient to consider the ball of radius $r_i \eps / (1 + \eps)$.
    In the figure this radius corresponds to the width of the annulus surrounding $q$,
    and it is  geometrically clear that this ball is indeed smaller than the one we could have discarded.
    The advantage of choosing this small radius is that, if we divide it by $2r_i$, we 
    get the expression $\eps / (2(1 + \eps))$, that depends only on $\eps$.
    By lemma \ref{lem:packing_lemma_dd} we know that there is an upper bound $T$ for the number of points
    $p_i$ inside the ball $B(p_i, 2r)$ that are at the distance at least $r_i / 2 ( 1 + \eps)$
    from each other, and the constant $T$ depends only on $\eps$ and the doubling dimension.
    This means that even in the worst case after we have computed $T$ distances, we have discarded all
    points of $P$ that lie outside of the ball $B(q, r_i)$, because all other points that lie in $B(p_i, 2r_i) \setminus B(q, r_i)$ 
    do not require distance computation. Indeed, each of these points lies in
    one of the balls $B(p_j, r_i \eps / (1 + \eps)$ for whose center we computed the distance already,
    and is therefore discarded, Now, either we have exhausted the set $P$, and in this
    case we terminate with an approximate nearest neighbor, or the next available $p_j$ that requires
    distance computation necessarily lies in $B(q, r_i)$, thus we found a closer neighbor. This means that $C = T+1$.
    Note that the algorithm still computes an approximate nearest neighbor, not the exact one, 
    because some of the points that are inside the annulus can be discarded, 
    if there is a point from $P$ close to them but outside of the annulus. 
\end{proof}

Now the theorem follows from the lemma by standard argument
of the backwards analysis.

>>>>

\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{comment}
\usepackage{braket}

@@

\begin{algorithmic}
\label{alg:ann_blind}
\Procedure{UpdateBounds}{$p_i, r_i$}
    \For{$k = i + 1, \dots, n$}
        \State {$a_k \gets \max(a_k, |\dist(p_i, p_k) - r_i|)$}
    \EndFor
\EndProcedure

\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
    \State {$[p_1, \dots, p_n] \gets \mbox{random permutation of }P$}
    \State { $a_i \gets 0$}
    \Comment {$a_i$ is lower bound for $\dist(p_i, q)$}
    \State {$ c \gets p_1, \quad v \gets \dist(p_1, q)$}
    \Comment {$c$ keeps the current candidate}
    \State \Call {UpdateBounds}{$p_1, v$}
    \For{$i = 2\dots n$}
        \If {$a_i \geq \frac{v}{1+\eps}$}
            \State {\textbf{continue}}
        \Else
            \State {Compute $r_i = \dist(p_i, q)$}
            \State \Call{UpdateBounds}{$p_i, r_i$}
            \If {$r_i < v$}
                \State {$c \gets p_i,\quad v\gets r_i$}
            \EndIf
        \EndIf
    \EndFor
    \State \Return {$c, v$}
\EndFunction
\end{algorithmic}

The algorithm that we analyze can be described as follows. 
Fix a random permutation of the points of $P$ (to simplify notation,
we re-index them, so the order is again $p_1, \dots, p_n$). Pick $p_1$
as the current candidate, denote $v = r_1$ and start traversing the list of $p_i$'s.
At each element $p_i$ look at the lower bound for $r_i$,
if it is larger than $v / (1+\eps)$, then we can safely skip $p_i$.
Otherwise we must compute the distance $r_i$, and we immediately
 use it to update the lower bounds for all the next points $p_k$'s. 
If $p_i$ turns out to be a better candidate (we compare $r_ir$ with $v$, not with $v/(1+\eps)$),
we update the current candidate $c$ and the distance to the query point $v$ accordingly, 
otherwise $c$ and $v$ remain unchanged.

@@

%\begin{algorithmic}
%\Procedure{Discard\_1}{$L, p_i$}
%    \State {$r_i = \dist(q, p_i)$}
%    \ForAll { $p_j \in L$}
%       \If {  $ \dist(p_i, p_j) > r_i \left(1 + \frac{1}{ 1 + \eps} \right)$ }
%           \State {Remove $p_j$ from $L$}
%       \EndIf
%    \EndFor
%\EndProcedure
%\end{algorithmic}

%\begin{algorithmic}
%\Procedure{Discard\_2}{$L, w, c$}
%    \State {$r_w = \dist(q, w)$}
%    \State {$r_c = \dist(q, w)$}
%    \Comment {Distances $r_w$ and $r_c$ are cached}
%    \ForAll { $p_j \in L$}
%       \If { $ \dist(w, p_j) < r_w - r_c / ( 1 + \eps) $ }
%           \State {Remove $p_j$ from $L$}
%           \Comment {In particular, the worse candidate $w$ will be removed here}
%       \EndIf
%       \If {  $ \dist(w, p_j) > r_w + r_c / ( 1 + \eps) $ }
%           \State {Remove $p_j$ from $L$}
%       \EndIf
%    \EndFor
%\EndProcedure
%\end{algorithmic}

%\begin{algorithmic}
%\label{alg:ann_blind}
%\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
%    \State {Initialize array $L$ with a random permutation of $P$}
%    \State {$ c \gets \mbox{the first element of }L$}
%    \State {Compute $r_i = \dist(q, p_i)$}
%    \State \Call{Discard\_1}{$L, c$}
%    \While {$|L| > 1$}
%        \ForAll {$p_j \in L $}
%            \State {Compute $r_j = \dist(p_j, q)$}
%            \Comment {Set the worse candidate $w$ to be either $p_i$ or $p_j$}
%            \If {$r_j < r_i$ }
%                \State {$ c \gets p_j $}
%                \State \Call{Discard\_1}{$L, c$}
%                \State {$ w \gets p_i $}
%            \Else
%                \State {$ w \gets p_i $}
%            \EndIf
%            \State \Call{Discard\_2}{$L, w, c$}
%        \EndFor
%    \EndWhile
%    \State \Return {the unique element left in $L$}
%\EndFunction
%\end{algorithmic}

@@

\begin{proof}
Let $i_1,...,i_k$ denote the indices of the minima in the sequence $r_1, \dots, r_n$
    (recall that this is a random sequence). Then $k=O(log n)$,
    by the standard backwards analysis argument.

    Consider two consecutive minima  $s:=i_j$ and $t:=i_{j+1}$. We claim that in the sequence
$p_s,...,p_{t-1}$
only a constant number of distances are computed, irrespective of its length.

To see that, let $c_1,...,c_n$ be the candidates after the $i$-th step of
the for loop of the algorithm. We have no guarantee that $c_s=p_s$, because $p_s$ it is possible
    that $p_s$ was discarded without computing $r_s$ (see the picture). However, we know that
$d(c_s,q)<=(1+eps) r_s$.

Indeed, suppose that $c_s \neq p_s$ and $r_s < d(c_s,q)/(1+eps)$. In this
    case $p_s$ cannot be discarded 
by its lower bound, and hence the distance to $p_s$ would
have been queried, but this distance is a minimum among $r_1, \dots, r_s$,
    and then we must reset $c_s$ to be $p_s$, a contradiction.

It follows that any point in the sequence $p_s,...,p_{t-1}$ for which a
distance is queried must lie in the ball around $c_s$ of radius at most $2(1+eps)r_s$.

Moreover, we know that for any index $j=s,\dots, t-1$ it holds
$d(c_j,q)>=r_s$.

Indeed, assume the opposite. $c_j$ is equal to $p_k$ for some $k<t$. If $k<s$,
that is, the candidate came from before, than $p_s$ would not be in the
minimum sequence. If $s<k<t$, then it follows that at least one of the
elements in $p_{s+1},\dots,p_k$ is in the minimum sequence, contradicting
the assumption that $p_s$ and $p_t$ are consecutive minima.

Hence, it follows that whenever a distance is queried, we can exclude a
ball of radius $\eps r_s$ around the query point. All queried points must
hence have pairwise distance of at least $ \eps r_s / 2$. Hence, the number of
queried points is by the packing lemma bounded by some constant that depends only on $\eps$ and the dimension. 
\end{proof}

%\begin{proof}
%    We start with the lemma of purely combinatorial (or probabilistic) nature.
%\begin{lemma}
%    Let $z$ be a random permutation of $n$ distinct real numbers,
%    $z = X_{1}, \dots, X_{n}$, and denote $x_t = \min \Set{X_1, \dots, X_t}$.
%    The expected number of positions $t$ where the sequence $X_t$ changes is $O(n)$.
%\end{lemma}
%    \begin{proof}
%        The proof is easily obtained with the standard backwards analysis technique and can be found in the appendix.
%        %Define random variable $Y_i$ to be 1, if $X_{i-1} = X_{i}$, and 0 otherwise.
%        %We must estimate $Y = \sum_{i = 2}^n Y_i$, and 
%        %\[
%        %    E[Y] = \sum_{i = 2}^n E[Y_i] = \sum_{i = 2}^n P(Y_i = 1).
%        %\]
%        %It remains to observe than $P(Y_i) = 1 / (n - i)$, so the expectation
%        %of $Y$ is the sum of the 
%    \end{proof}

%\begin{lemma}
%    There exists a constant $C$ that depends only on $\eps$ and the doubling dimension
%    such that after computing at most $C$ distances $\dist(p_k, q)$
%    either the algorithm verifies that the current candidate $p_i$ is a correct answer and terminates
%    or the algorithm finds another candidate $p_j$ such that $r_j < r_i$.
%\end{lemma}
%\begin{proof}
%    First, all points that are not contained in the ball $B(p_i, 2r_i)$
%    are discarded without computing the distance to them.
%    Now we have to examine all points $p_j$ inside this ball. 
%    Suppose that we cannot discard the current point $p_j$ based
%    on the lower bound for $r_j$, then we must compute the distance $r_j$.
%    There are two possibilities: either $r_j \leq r_i$ or $r_j > r_i$.
%    The latter case means that we discard the point $p_j$,
%    but we also discard all points of $P$ that are contained in some ball around it.
%    The triangle inequality tells us that we can take the ball of radius $r_j - r_i / (1 + \eps)$,
%    but in this proof it is sufficient to consider the ball of radius $r_i \eps / (1 + \eps)$.
%    In the figure this radius corresponds to the width of the annulus surrounding $q$,
%    and it is  geometrically clear that this ball is indeed smaller than the one we could have discarded.
%    The advantage of choosing this small radius is that, if we divide it by $2r_i$, we 
%    get the expression $\eps / (2(1 + \eps))$, that depends only on $\eps$.
%    By lemma \ref{lem:packing_lemma_dd} we know that there is an upper bound $T$ for the number of points
%    $p_i$ inside the ball $B(p_i, 2r)$ that are at the distance at least $r_i / 2 ( 1 + \eps)$
%    from each other, and the constant $T$ depends only on $\eps$ and the doubling dimension.
%    This means that even in the worst case after we have computed $T$ distances, we have discarded all
%    points of $P$ that lie outside of the ball $B(q, r_i)$, because all other points that lie in $B(p_i, 2r_i) \setminus B(q, r_i)$ 
%    do not require distance computation. Indeed, each of these points lies in
%    one of the balls $B(p_j, r_i \eps / (1 + \eps)$ for whose center we computed the distance already,
%    and is therefore discarded, Now, either we have exhausted the set $P$, and in this
%    case we terminate with an approximate nearest neighbor, or the next available $p_j$ that requires
%    distance computation necessarily lies in $B(q, r_i)$, thus we found a closer neighbor. This means that $C = T+1$.
%    Note that the algorithm still computes an approximate nearest neighbor, not the exact one, 
%    because some of the points that are inside the annulus can be discarded, 
%    if there is a point from $P$ close to them but outside of the annulus. 
%\end{proof}

%Now the theorem follows from the lemma by standard argument
%of the backwards analysis.

%\end{proof}
